---
title: "Data Preparation and Exploration"
output: html_notebook
---

# Approach (to be defined)


```{r Libraries, message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(magrittr)
library(DataExplorer)
library(plotly)
library(skimr)
```


# Data 
Practice level prescribing data is a list of all medicines, dressings and appliances that are prescribed by all practices in England, including GP practices, each month.

Data understanding:

- [Original source of data at NHS and essential information](https://digital.nhs.uk/data-and-information/areas-of-interest/prescribing/practice-level-prescribing-in-england-a-summary) (28. March 2019)
- [Glossary of terms](https://digital.nhs.uk/data-and-information/areas-of-interest/prescribing/practice-level-prescribing-in-england-a-summary/practice-level-prescribing-glossary-of-terms) (28. March 2019) 
- [BNF Codes](https://ebmdatalab.net/prescribing-data-bnf-codes/) (28. March 2019) 

## Files
For each month there are three files:
- Practice Prescribing Data file - this is the main data file giving the practice level prescribing data at presentation level using the full 15 digit BNF code. 
- GP prescribing chemical substance file - this gives the chemical substance or section as appropriate using the 9 digit (Drugs) or 4 digit (Appliances) BNF code. 
- GP prescribing practice address file - this gives the practice name and address. 


## Read files

Problems we solved:
- which is the data model regarding the 3 files? 
- which are the headers for file T201807ADDR BNFT.CSV? 
- check columns 3 & 4 in T201807CHEM SUBS.CSV 
- lowercase all variables and combine space separated variables with _  
- check last column in T201807PDPI BNFT.CSV and deal with it in an appropriate way 


```{r Read Files, eval=FALSE, include=FALSE}
gp <- read_csv(file = "data_files_2018_07_Jul/T201807ADDR BNFT.CSV", col_names = F)
# col_names = c("date_month", "tbd", "practioner_1", "practioner_2") total 8
# lower all coluns
names(gp) %<>% tolower

# rename variables with suitable headers
names(gp) = c("period", "practice_code","practice_name","adress_1",
                     "adress_2","adress_3","adress_4","postcode")


chem_subs <- read_csv(file = "data_files_2018_07_Jul/T201807CHEM SUBS.CSV")
head(chem_subs)
chem_subs$`chem sub`
names(chem_subs) %<>% tolower
chem_subs %<>% select(-x4)
chem_subs %<>%
  mutate(`201807` = coalesce(`201807`, "201807")) %>% 
  select(year_month = `201807`, chem_sub = `chem sub`, everything())


# plpd = Practice level prescribing data
plpd <- read_csv(file = "data_files_2018_07_Jul/T201807PDPI BNFT.CSV")
head(plpd)
names(plpd) %<>% tolower


# check last column: it is just an empty column
read_lines(file = "data_files_2018_07_Jul/T201807PDPI BNFT.CSV", n_max = 10)

# drop empty column x11
plpd %<>% select(-x11)


# rename variables containing ''
plpd %<>%
  select(bnf_code = `bnf code`, bnf_name = `bnf name`, act_cost = `act cost`, everything())

# preparation inputs from Philipp Hochueli 
## drop non proper BNF-codes (less than 15 characters)
plpd_smaller %<>% 
  mutate(bnf_code_length = nchar(bnf_code)) %>% 
  filter(bnf_code_length < 15) %>% 
  select(-bnf_code_length)

## drop actual costs & quantity
plpd %<>% select(-c(act_cost, quantity))

# cast to right datatype
plpd$nic <- as.double(plpd$nic)
plpd$items <- as.integer(plpd$items)

# generate 5% plpd sample regarding computing efficiency
set.seed(22)
plpd_sample <- sample_frac(plpd, 0.05)
summary(plpd_sample)
glimpse(plpd_sample) 

rm(dlpd)
```

# Data Model
Possible solution:

- extract chemical substance from bnft table from the attribute "bnf_name". connect the chemical substance from the bnft with the chem_subs file (attribute "chem_sub"). Assumption: mapping with the first 9 signs, example: 040702040 
- *what about the  4 digit (Appliances) BNF code?* 
- Attribute "practice" from bnft table can be connected with the second attribute (name tbd) from the adrr table, example: B86057 

```{r Data Model, eval=FALSE, include=FALSE}
# extract chem_sub in plpd
plpd_sample %<>% 
  mutate(chem_sub = str_sub(bnf_code, 1, 9))

# join plpd with gp & chem_subs
d <- left_join(x = plpd_sample, y = gp, by = c("practice" = "practice_code"), suffix = c(".plpd", ".gp")) %>%
  left_join(chem_subs, by = "chem_sub") %>% 
  select(bnf_code, bnf_name, chem_sub_code = chem_sub, chem_name = name, items, nic, 
         practice_code = practice, practice_name, sha, adress_1,
         adress_2, adress_3, adress_4, postcode, period = period.plpd)
# not including pct because sha is better according to Philipp Hochueli

# split bnf_name into drug_dressing_appliances, forms & dosages with tidyr
d %<>%
  separate(bnf_name, into = c("drug_dressing_appliances", "form"),
           sep = "_", remove = FALSE) %>% 
  separate(form, into = c("form", "dosage"),
           sep = " ", remove = TRUE, extra = "merge") 
# check error message: Expected 2 pieces. Missing pieces filled with `NA` in 10 rows [22980, 30494, 110893, 213266, 270869, 305394, 312425, 341193, 360483, 366071].Expected 2 pieces. Missing pieces filled with `NA` in 25704 rows [23, 53, 55, 112, 150, 165, 185, 213, 229, 246, 271, 276, 307, 326, 426, 440, 457, 519, 526, 553, ...].

d %<>% 
  mutate(nic_item = nic/items)

# write_csv(x = d, path = "plpd_extended.csv")


```


```{r Load sample data directly, message=FALSE, warning=FALSE}
# load preprocessed data
d <- read_csv("plpd_extended.csv")
```


# Data Exploration
```{r Data Exploration}
glimpse(d)

# explore data with DataExplorer package
introduce(d)
plot_intro(d)
plot_missing(d)

# explore numerical values
# create_report(d[, c("items", "nic")])

```




# Scope 

## What we can't answer with the given data

You can't:

- download data for only one practice or subset of practices - the data is contained in a single large data file 
- compare one practice with another without further information about prescribing, list sizes, age and gender of patients on the list, knowledge about the medicines and why they are prescribed 
- identify individual patients, even those receiving medication for rare conditions, because no patient data is contained in the data 
- find out how many people with a particular condition have been given prescriptions - prescriptions given to patients do not say why the drug has been prescribed 
- find out which pharmacies dispensed these prescriptions 
- find out about private prescrptions - these are not included in the data 
- find out the proportion of prescriptions that were paid for by patients or exempt from payment 

With this data set alone it is not possible to work out the prescribing rate - the percentage of patients receiving prescriptions. You would need additional data on the number of patients registered with a practice which can be found on the NHS Business Services Authority website. 

[Source](https://digital.nhs.uk/data-and-information/areas-of-interest/prescribing/practice-level-prescribing-in-england-a-summary/practice-level-prescribing-data-more-information) (04.04.2019). 

## Questions to be answered 
*to be added*

# Analysis

## Merge Branded drugs & generics


```{r Branded drugs & generics}
# create a DF that contains every generic drug just once
generic_list <- d %>%
  filter(str_sub(bnf_code,10,11) == 'AA') %>% #drugs with no 'AA' at bnf letter 10 and 11 branded (no generics)
  group_by(bnf_code, bnf_name, drug_dressing_appliances, form, dosage, chem_name) %>%
  summarise(nic_sum=sum(nic), items_sum=sum(items)) %>%
  ungroup()
generic_list <- generic_list[order(generic_list$items_sum,decreasing=TRUE),] # order list for 'item_sum'
# create a key for later join with the branded drug
generic_list %<>% 
  mutate(bnf_key = paste(str_sub(bnf_code, 1, 9), str_sub(bnf_code,12,13), sep="")) %>% 
  mutate(nic_per_item = nic_sum/items_sum) # calculate nic per item
generic_list <- generic_list[c(9,1,2,3,4,5,6,7,8,10)]

# create a DF that contains every original drug just once
branded_list <- d %>%
  filter(str_sub(bnf_code,10,11) != 'AA') %>% #drugs with 'AA' at bnf letter 10 and 11 are generics
  group_by(bnf_code, bnf_name, drug_dressing_appliances, form, dosage, chem_name) %>%
  summarise(nic_sum=sum(nic), items_sum=sum(items)) %>%
  ungroup()
branded_list <- branded_list[order(branded_list$items_sum,decreasing=TRUE),]
# create a key for later join with the generic drug
branded_list %<>% 
  mutate(bnf_key = paste(str_sub(bnf_code, 1, 9), str_sub(bnf_code,14,15), sep="")) %>%
  mutate(nic_per_item = nic_sum/items_sum) # calculate nic per item
branded_list <- branded_list[c(9,1,2,3,4,5,6,7,8,10)]

complete_list <- d %>%
  group_by(bnf_code) %>%
  summarise(nic_sum=sum(nic), items_sum=sum(items)) %>%
  ungroup()

# left join for every branded drug the generic drug next to it. ->  verify if there is a drug with two or more generics
med_list <- left_join(x = branded_list, y = generic_list, by = c("bnf_key"), suffix = c("_branded", "_generic"))
med_list_a <- med_list %>% select(bnf_code_branded, bnf_name_branded, bnf_code_generic, bnf_name_generic)


# plausability check
# number of generics in 'generic_list': 3537
# number of generics in 'med_list': 
# number of unique generics in 'med_list: 


# Which drug is more expensive?
orginal_expensiv <- med_list %>%
  filter(med_list$nic_per_item_branded > med_list$nic_per_item_generic)
generic_expensiv <- med_list %>%
  filter(med_list$nic_per_item_branded < med_list$nic_per_item_generic)

boxplot(med_list$nic_per_item_branded, med_list$nic_per_item_generic, 
        names=c("Branded","Generic"))

# explore data with DataExplorer package
introduce(branded_list)
plot_intro(branded_list)
plot_missing(branded_list, title = "Branded")

introduce(generic_list)
plot_intro(generic_list)
plot_missing(generic_list, title = "Generics")

attach(med_list)
# inspect correlation and N/A values
# create_report(med_list[, c("nic_sum_branded", "nic_sum_generic", "items_sum_branded", "items_sum_generic", "nic_per_item_branded", "nic_per_item_generic")])
plot_missing(med_list)
plot_histogram(med_list)

# data_representation_for_slides <- med_list %>% 
#   select(1:2, 8:10, 11:12, 17:19)
```




## Questions: How many branded drugs have an equal generic? How many branded products do not have a generic?

```{r Question x}
# Wie viele Branded-Products haben ein Generika? Wie viele Branded-Products haben kein Generika?
skim(med_list)
(med_list_missing_values <- as.factor(sapply(med_list, function(x) sum(is.na(x)))))
(missing_ration <- as.numeric(levels(med_list_missing_values)[med_list_missing_values])[11]/
  length(bnf_code_branded))

# @Lars: how many generica don't have a fitting branded drug?
```

Answer: There are 1260 missing values for the generics data. There are totally 4142 in our joined branded drugs and generics table. The missing ratio is therefore about 30%. Conclusion: There are close to *70% (about 2900) of all branded drugs with a generic and 30% (about 1242) vice versa*.




## Question: How is the price difference between branded drugs & generics?

Approach: Compare the branded drugs prices (sum & per item) with the generic prices (sum & per item) as well as the sold items sum of both of them. 


```{r Question xx, warning=FALSE}
# Wie ist der Preisunterschied zwischen Branded-Products und Generikas? Sind Generikas billiger?

# comparisons branded drugs vs. generic
## add price per item difference
med_list %<>% 
  mutate(diff_branded_to_generic = nic_per_item_branded - nic_per_item_generic)

attach(med_list)


## show distribution comparison price sum
(p_nic_sum <- plot_ly(x = ~generic_list$nic_sum, name = "Generic price sum", type = "box", boxpoints = "all", jitter = 0.25) %>%
  add_trace(x = ~branded_list$nic_sum, name = 'Branded price sum') %>%
  layout(title = "Price sum comparison"))
## compare mean
summary(generic_list$nic_sum)
summary(branded_list$nic_sum)


## show distribution comparison price per item 
(p_nic_per_item <- plot_ly(x = ~nic_per_item_generic, name = "Generic price per item", type = "box", boxpoints = "all", jitter = 0.25) %>%
  add_trace(x = ~nic_per_item_branded, name = 'Branded price per item') %>%
  layout(title = "Price per item comparison"))


## show distribution comparison difference price per item
(
  p_nic_per_item_difference_boxplot <-
  plot_ly(
  x = ~diff_branded_to_generic,
  type = "box",
  name = " ",
  boxpoints = "all",
  jitter = 0.25
  ) %>%
  layout(title = "Item price branded drug - generic price")
  )

summary(diff_branded_to_generic)

(p_nic_per_item_difference_hist <- plot_ly(x = ~diff_branded_to_generic, type = "histogram", histnorm = "probability", alpha = 0.5, nbinsx = 2000) %>% 
  layout(title = "Frequency item price branded drug - generic price",
         xaxis = list(title = "item price branded drug - generic price",
                      zeroline = FALSE),
         yaxis = list(title = "Frequency",
                      zeroline = FALSE)))



## show distribution comparison items sold sum total 
(p_item_sum <- plot_ly(x = ~generic_list$items_sum, name = "Generic sold items sum", type = "box", boxpoints = "all", jitter = 0.25) %>%
  add_trace(x = ~branded_list$items_sum, name = 'Branded sold items sum') %>%
  layout(title = "Item sold sum comparison"))
## compare mean
summary(generic_list$items_sum)
summary(branded_list$items_sum)

detach(med_list)
```

Answers: 

- The total price of the sold generics is much higher for generics (median = ~833) than for branded drugs (median = ~265). The generics price sum variates much more than the branded drugs one. This numbers are caused by the totally more sold generics (65% of all items sold). 
- The price per item comparison shows that the branded drugs are overall 0.18 pounds more expensive (median). The first quartile tells us that 25% of all comparisons are at least 2.71 pound cheaper (generics vs. branded drugs). There are on the other side 25% of all comparisons at least 4.82 pound more expensive (branded drugs vs. generics). The mean is much different with ~3.1 pound more expensive branded drugs because it is influenced a lot by outliers. That's why the median is more suitable in this case. *Outlook: one should compare prices with classes, for example by drug chapters*. 
- Generics represent total sold items of 35 (median) vs. branded drugs with 12 (median). This means there are in average about three times more sold generics then branded drugs! The total sold items of generics  spread much more than the branded drugs. The mean confirms the discovery: Generics are in average sold 1041 times versus 168 times (generic drugs). There are a lot of outliers in this case, too. Therefore it is reasonable to stay with the median.    








