---
title: "Data Preparation and Exploration"
output: html_notebook
---

# Approach (to be defined)


```{r Libraries, message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(magrittr)
library(DataExplorer)
```


# Data 
Data understanding:

- [Original source of data at NHS and essential information](https://digital.nhs.uk/data-and-information/areas-of-interest/prescribing/practice-level-prescribing-in-england-a-summary) (28. March 2019)
- [Glossary of terms](https://digital.nhs.uk/data-and-information/areas-of-interest/prescribing/practice-level-prescribing-in-england-a-summary/practice-level-prescribing-glossary-of-terms) (28. March 2019) 
- [BNF Codes](https://ebmdatalab.net/prescribing-data-bnf-codes/) (28. March 2019) 

Practice level prescribing data is a list of all medicines, dressings and appliances that are prescribed by all practices in England, including GP practices, each month.

## Files
For each month there are three files:
- Practice Prescribing Data file - this is the main data file giving the practice level prescribing data at presentation level using the full 15 digit BNF code. 
- GP prescribing chemical substance file - this gives the chemical substance or section as appropriate using the 9 digit (Drugs) or 4 digit (Appliances) BNF code. 
- GP prescribing practice address file - this gives the practice name and address. 


## Read files

Problems to solve:
- which is the data model r♣egarding the 3 files? 
- which are the headers for file T201807ADDR BNFT.CSV 
- check columns 3 & 4 in T201807CHEM SUBS.CSV 
- lowercase all variables and combine space separated variables with _  
- check last column in T201807PDPI BNFT.CSV 


```{r Read Files}
gp <- read_csv(file = "data_files_2018_07_Jul/T201807ADDR BNFT.CSV", col_names = F)
# col_names = c("date_month", "tbd", "practioner_1", "practioner_2") total 8
# lower all coluns
names(gp) %<>% tolower

# rename variables with suitable headers
names(gp) = c("period", "practice_code","practice_name","adress_1",
                     "adress_2","adress_3","adress_4","postcode")


chem_subs <- read_csv(file = "data_files_2018_07_Jul/T201807CHEM SUBS.CSV")
head(chem_subs)
chem_subs$`chem sub`
names(chem_subs) %<>% tolower
chem_subs %<>% select(-x4)
chem_subs %<>%
  mutate(`201807` = coalesce(`201807`, "201807")) %>% 
  select(year_month = `201807`, chem_sub = `chem sub`, everything())


# plpd = Practice level prescribing data
plpd <- read_csv(file = "data_files_2018_07_Jul/T201807PDPI BNFT.CSV")
head(plpd)
names(plpd) %<>% tolower


# check last column: it is just an empty column
read_lines(file = "data_files_2018_07_Jul/T201807PDPI BNFT.CSV", n_max = 10)

# drop empty column x11
plpd %<>% select(-x11)


# rename variables containing ''
plpd %<>%
  select(bnf_code = `bnf code`, bnf_name = `bnf name`, act_cost = `act cost`, everything())

# preparation inputs from Philipp Hochueli 
## drop non proper BNF-codes (less than 15 characters)
plpd_smaller %<>% 
  mutate(bnf_code_length = nchar(bnf_code)) %>% 
  filter(bnf_code_length < 15) %>% 
  select(-bnf_code_length)

## drop actual costs & quantity
plpd %<>% select(-c(act_cost, quantity))

# cast to right datatype
plpd$nic <- as.double(plpd$nic)
plpd$items <- as.integer(plpd$items)

# generate 5% plpd sample regarding computing efficiency
set.seed(22)
plpd_sample <- sample_frac(plpd, 0.05)
summary(plpd_sample)
glimpse(plpd_sample) 

rm(dlpd)

```

# Data Model
Possible solution:

- extract chemical substance from bnft table from the attribute "bnf_name". connect the chemical substance from the bnft with the chem_subs file (attribute "chem_sub"). Assumption: mapping with the first 9 signs, example: 040702040 
- *what about the  4 digit (Appliances) BNF code?* 
- Attribute "practice" from bnft table can be connected with the second attribute (name tbd) from the adrr table, example: B86057 

```{r Data Model}
# extract chem_sub in plpd
plpd_sample %<>% 
  mutate(chem_sub = str_sub(bnf_code, 1, 9))

# join plpd with gp & chem_subs
d <- left_join(x = plpd_sample, y = gp, by = c("practice" = "practice_code"), suffix = c(".plpd", ".gp")) %>%
  left_join(chem_subs, by = "chem_sub") %>% 
  select(bnf_code, bnf_name, chem_sub_code = chem_sub, chem_name = name, items, nic, 
         practice_code = practice, practice_name, sha, adress_1,
         adress_2, adress_3, adress_4, postcode, period = period.plpd)
# not including pct because sha is better according to Philipp Hochueli

# split bnf_name into drug_dressing_appliances, forms & dosages with tidyr
d %<>%
  separate(bnf_name, into = c("drug_dressing_appliances", "form"),
           sep = "_", remove = FALSE) %>% 
  separate(form, into = c("form", "dosage"),
           sep = " ", remove = TRUE, extra = "merge") 
# check error message: Expected 2 pieces. Missing pieces filled with `NA` in 10 rows [22980, 30494, 110893, 213266, 270869, 305394, 312425, 341193, 360483, 366071].Expected 2 pieces. Missing pieces filled with `NA` in 25704 rows [23, 53, 55, 112, 150, 165, 185, 213, 229, 246, 271, 276, 307, 326, 426, 440, 457, 519, 526, 553, ...].

d %<>% 
  mutate(nic_item = nic/items)

# write_csv(x = d, path = "plpd_extended.csv")


```


```{r Load sample data directly}
# load preprocessed data
d <- read_csv("plpd_extended.csv")
```


# Data Exploration
```{r Data Exploration}
glimpse(d)

# explore data with DataExplorer package
introduce(d)
plot_intro(d)
plot_missing(d)

# explore numerical values
# create_report(d[, c("items", "nic")])

```




# Scope 

## What we can't answer with the given data

You can't:

- download data for only one practice or subset of practices - the data is contained in a single large data file 
- compare one practice with another without further information about prescribing, list sizes, age and gender of patients on the list, knowledge about the medicines and why they are prescribed 
- identify individual patients, even those receiving medication for rare conditions, because no patient data is contained in the data 
- find out how many people with a particular condition have been given prescriptions - prescriptions given to patients do not say why the drug has been prescribed 
- find out which pharmacies dispensed these prescriptions 
- find out about private prescrptions - these are not included in the data 
- find out the proportion of prescriptions that were paid for by patients or exempt from payment 

With this data set alone it is not possible to work out the prescribing rate - the percentage of patients receiving prescriptions. You would need additional data on the number of patients registered with a practice which can be found on the NHS Business Services Authority website. 

[Source](https://digital.nhs.uk/data-and-information/areas-of-interest/prescribing/practice-level-prescribing-in-england-a-summary/practice-level-prescribing-data-more-information) (04.04.2019). 


```{r Branded drugs & Generics}
# Code Lars


# create a DF that contains every generic drug just once
generic_list <- d %>%
  filter(str_sub(bnf_code,10,11) == 'AA') %>% #drugs with no 'AA' at bnf letter 10 and 11 branded (no generics)
  group_by(bnf_code, bnf_name, drug_dressing_appliances, form, dosage, chem_name) %>%
  summarise(nic_sum=sum(nic), items_sum=sum(items)) %>%
  ungroup()
generic_list <- generic_list[order(generic_list$items_sum,decreasing=TRUE),] # order list for 'item_sum'
# create a key for later join with the branded drug
generic_list %<>% 
  mutate(bnf_key = paste(str_sub(bnf_code, 1, 9), str_sub(bnf_code,12,13), sep="")) %>% 
  mutate(nic_per_item = nic_sum/items_sum) # calculate nic per item
generic_list <- generic_list[c(9,1,2,3,4,5,6,7,8,10)]

# create a DF that contains every original drug just once
branded_list <- d %>%
  filter(str_sub(bnf_code,10,11) != 'AA') %>% #drugs with 'AA' at bnf letter 10 and 11 are generics
  group_by(bnf_code, bnf_name, drug_dressing_appliances, form, dosage, chem_name) %>%
  summarise(nic_sum=sum(nic), items_sum=sum(items)) %>%
  ungroup()
branded_list <- branded_list[order(branded_list$items_sum,decreasing=TRUE),]
# create a key for later join with the generic drug
branded_list %<>% 
  mutate(bnf_key = paste(str_sub(bnf_code, 1, 9), str_sub(bnf_code,14,15), sep="")) %>%
  mutate(nic_per_item = nic_sum/items_sum) # calculate nic per item
branded_list <- branded_list[c(9,1,2,3,4,5,6,7,8,10)]

complete_list <- d %>%
  group_by(bnf_code) %>%
  summarise(nic_sum=sum(nic), items_sum=sum(items)) %>%
  ungroup()

# left join for every branded drug the generic drug next to it. ->  verify if there is a drug with two or more generics
med_list <- left_join(x = branded_list, y = generic_list, by = c("bnf_key"), suffix = c("_branded", "_generic"))
med_list_a <- med_list %>% select(bnf_code_branded, bnf_name_branded, bnf_code_generic, bnf_name_generic)


# plausability check
# number of generics in 'generic_list': 3537
# number of generics in 'med_list': 
# number of unique generics in 'med_list: 


# Which drug is more expensive?
orginal_expensiv <- med_list %>%
  filter(med_list$nic_per_item_branded > med_list$nic_per_item_generic)
generic_expensiv <- med_list %>%
  filter(med_list$nic_per_item_branded < med_list$nic_per_item_generic)

boxplot(med_list$nic_per_item_branded, med_list$nic_per_item_generic, 
        names=c("Branded","Generic"))

# explore data with DataExplorer package
introduce(branded_list)
plot_intro(branded_list)
plot_missing(branded_list, title = "Branded")

introduce(generic_list)
plot_intro(generic_list)
plot_missing(generic_list, title = "Generics")

attach(med_list)
# inspect correlation and N/A values
create_report(med_list[, c("nic_sum_branded", "nic_sum_generic", "items_sum_branded", "items_sum_generic", "nic_per_item_branded", "nic_per_item_generic")])

```

```{r Question x}
# Wie viele Branded-Products haben ein Generika? Wie viele Branded-Products haben kein Generika?
plot_ly(x = ~items_sum.generic, name = "Generic price sum", type = "box", boxpoints = "all", jitter = 0.25)




```


```{r Question xx}
# Wie ist der Preisunterschied zwischen Branded-Products und Generikas? Sind Generikas billiger?
library(plotly)

# comparisons branded vs. generic
## add price per item difference
med_list %>% 
  mutate(diff_branded_to_generic = nic_per_item.original - nic_per_item.generic)

attach(med_list)

## show distribution comparison price sum
(p_nic_sum <- plot_ly(x = ~nic_sum.generic, name = "Generic price sum", type = "box", boxpoints = "all", jitter = 0.25) %>%
  add_trace(x = ~nic_sum.original, name = 'Original price sum'))


## show distribution comparison price per item 
(p_nic_per_item <- plot_ly(x = ~nic_per_item.generic, name = "Generic price per item", type = "box", boxpoints = "all", jitter = 0.25) %>%
  add_trace(x = ~nic_per_item.original, name = 'Original price per item'))

## calculate price per item difference 



## show distribution comparison sum total 
(p_items_sum <- plot_ly(x = ~items_sum.generic, name = "Generic price sum", type = "box", boxpoints = "all", jitter = 0.25) %>%
  add_trace(x = ~items_sum.original, name = 'Original price sum'))


```




